Simple repression draft

### 2020-05-23 doc/appendix_03_regulated_promoters.tex
Text on posterior inference on repressed data

\label{eq:p_m_bursty+rep}

@@

\subsubsection{Moments}
We can verify the correctness of~\eq{eq:p_m_bursty+rep}
by comparison with Gillespie simulation.
We can also easily compute moments of the distribution from
the generating function, and check these against raw moments computed
directly from the master equation.

@@

\subsection{Numerical considerations and recursion formulas}
\subsubsection{Generalities}
We would like to carry out Bayesian parameter inference on FISH data
from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep} as our
likelihood. This requires accurate (and preferably fast)
numerical evaluation of the hypergeometric function ${_2F_1}$,
which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
and our particular needs here present an especial challenge as we show below.

The hypergeometric function is defined by its Taylor series as
\begin{align}
{_2F_1}(a,b,c;z) 
= \sum_{l=0}^\infty
\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
\frac{z^l}{l!}
\end{align}
for $|z|<1$, and by analytic continuation elsewhere.
If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
(absolute value below 20 or 30),
then the series converges quickly and an accurate numerical representation is
easily computed by truncating the series after a reasonable number of terms.
Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
to the tail of the distribution, which can easily reach 50, possibly 100.
From~\eq{eq:p_m_bursty+rep}, this means evaluating ${_2F_1}$
repeatedly for values of $a$, $b$, and $c$ spanning the full range
from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
even if $\alpha$, $\beta$, and $\gamma$
in~\eq{eq:p_m_bursty+rep} are small,
with the situation even worse if they are not small.
A naive numerical evaluation of the series definition will be
prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
series have alternating signs which can lead to catastrophic cancellations.

One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
instead of floating point arithmetic,
e.g., using the \texttt{mpmath} library in Python.
This is accurate but incredibly slow computationally.
To quantify how slow, we found that
evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep} $\sim50$ times
(for a typical dataset of interest from~\cite{Jones2014},
with $m$ values spanning 0 to $\sim50$)
using arbitrary precision arithmetic is 100-1000 fold slower than
evaluating a negative binomial likelihood for the corresponding
constitutive promoter dataset.

@@

\subsubsection{Particulars}
Let us now demonstrate these generalities for our problem of interest.
In order to evaluate the probability distribution of our
model,~\eq{eq:p_m_bursty+rep}, we need to evaluate hypergeometric functions
of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
The authors of~\cite{Gil2007} did not derive a recursion relation
for precisely this case. We could follow their methods and do so ourselves,
but it is much easier to convert to a case that they did consider.
The strategy is to look through the minimal solutions tabulated
in~\cite{Gil2007} and search for a transformation we could apply to
${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
(the variable being incremented by the recursion)
in the same arguments of ${_2F_1}$ as the minimal solution.
After some ``guess and check,'' we found that the transformation
\begin{align}
{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
=
(1+b)^{-\alpha-m}
        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
\label{eq:rec_euler_pretransform}
\end{align}
produces a ${_2F_1}$ on the right hand side that closely resembles
the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
Explicitly, these solutions are
\begin{align}
y_{3,m}
&\propto
{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
                -\beta^\prime + \gamma^\prime,
                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
                1-z\right)
\\
y_{4,m}
&\propto
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
                1-z\right),
\label{eq:minimal_soln_sans_prefac}
\end{align}
where we have omitted prefactors which are unimportant for now.
Which of these two we should use depends on what values $z$ takes on.
Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
$z$ is bounded between 0 and 1.
From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.

@@

With ${_2F_1}$ computed, the only remaining numerical danger in computing
$p(m)$ in~\eq{eq:p_m_bursty+rep} is overflow of the gamma functions.
This is easily solved by taking the log of the entire expression
and using standard routines to compute the log of the gamma functions,
then exponentiating the entire expression at the end if $p(m)$
is needed rather than $\log p(m)$.

>>>>

\label{eq:p_m_bursty+rep_appdx}

@@

\subsubsection{Moments}
We can verify the correctness of~\eq{eq:p_m_bursty+rep_appdx}
by comparison with Gillespie simulation.
We can also easily compute moments of the distribution from
the generating function, and check these against raw moments computed
directly from the master equation.

@@

\subsection{Numerical considerations and recursion formulas}
\subsubsection{Generalities}
We would like to carry out Bayesian parameter inference on FISH data
from~\cite{Jones2014}, using~\eq{eq:p_m_bursty+rep_appdx} as our
likelihood. This requires accurate (and preferably fast)
numerical evaluation of the hypergeometric function ${_2F_1}$,
which is a notoriously hard problem~\cite{Pearson2017, Gil2007},
and our particular needs here present an especial challenge as we show below.

The hypergeometric function is defined by its Taylor series as
\begin{align}
{_2F_1}(a,b,c;z) 
= \sum_{l=0}^\infty
\frac{\Gamma(a + l)\Gamma(b + l)\Gamma(c)}
        {\Gamma(a)\Gamma(b)\Gamma(c + l)}
\frac{z^l}{l!}
\end{align}
for $|z|<1$, and by analytic continuation elsewhere.
If $z\lesssim1/2$ and $\alpha$ and $\beta$ are not too large
(absolute value below 20 or 30),
then the series converges quickly and an accurate numerical representation is
easily computed by truncating the series after a reasonable number of terms.
Unfortunately, we need to evaluate ${_2F_1}$ over mRNA copy numbers fully out
to the tail of the distribution, which can easily reach 50, possibly 100.
From~\eq{eq:p_m_bursty+rep_appdx}, this means evaluating ${_2F_1}$
repeatedly for values of $a$, $b$, and $c$ spanning the full range
from $\mathcal{O}(1)$ to $\mathcal{O}(10^2)$,
even if $\alpha$, $\beta$, and $\gamma$
in~\eq{eq:p_m_bursty+rep_appdx} are small,
with the situation even worse if they are not small.
A naive numerical evaluation of the series definition will be
prone to overflow and, if any of $a,b,c<0$, then some successive terms in the
series have alternating signs which can lead to catastrophic cancellations.

One solution is to evaluate ${_2F_1}$ using arbitrary precision arithmetic
instead of floating point arithmetic,
e.g., using the \texttt{mpmath} library in Python.
This is accurate but incredibly slow computationally.
To quantify how slow, we found that
evaluating the likelihood defined by~\eq{eq:p_m_bursty+rep_appdx} $\sim50$ times
(for a typical dataset of interest from~\cite{Jones2014},
with $m$ values spanning 0 to $\sim50$)
using arbitrary precision arithmetic is 100-1000 fold slower than
evaluating a negative binomial likelihood for the corresponding
constitutive promoter dataset.

@@

\subsubsection{Particulars}
Let us now demonstrate these generalities for our problem of interest.
In order to evaluate the probability distribution of our
model,~\eq{eq:p_m_bursty+rep_appdx}, we need to evaluate hypergeometric functions
of the form ${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$
for values of $m$ ranging from $0$ to $\mathcal{O}(100)$.
The authors of~\cite{Gil2007} did not derive a recursion relation
for precisely this case. We could follow their methods and do so ourselves,
but it is much easier to convert to a case that they did consider.
The strategy is to look through the minimal solutions tabulated
in~\cite{Gil2007} and search for a transformation we could apply to
${_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)$ that would place the $m$'s
(the variable being incremented by the recursion)
in the same arguments of ${_2F_1}$ as the minimal solution.
After some ``guess and check,'' we found that the transformation
\begin{align}
{_2F_1}(\alpha+m, \beta+m, \gamma+m; -b)
=
(1+b)^{-\alpha-m}
        {_2F_1}\left(\alpha+m, \gamma-\beta, \gamma+m; \frac{b}{1+b}\right),
\label{eq:rec_euler_pretransform}
\end{align}
produces a ${_2F_1}$ on the right hand side that closely resembles
the minimal solutions $y_{3,m}$ and $y_{4,m}$ in Eq.~4.3 in~\cite{Gil2007}.
Explicitly, these solutions are
\begin{align}
y_{3,m}
&\propto
{_2F_1}\left(-\alpha^\prime + \gamma^\prime - m,
                -\beta^\prime + \gamma^\prime,
                1-\alpha^\prime-\beta^\prime+\gamma^\prime-m;
                1-z\right)
\\
y_{4,m}
&\propto
{_2F_1}\left(\alpha^\prime + m,
                \beta^\prime,
                1+\alpha^\prime+\beta^\prime-\gamma^\prime+m;
                1-z\right),
\label{eq:minimal_soln_sans_prefac}
\end{align}
where we have omitted prefactors which are unimportant for now.
Which of these two we should use depends on what values $z$ takes on.
Equating $1-z=b/(1+b)$ gives $z=1/(1+b)$, and since $b$ is strictly positive,
$z$ is bounded between 0 and 1.
From Eq.~4.5 in~\cite{Gil2007}, $y_{4,m}$ is the minimal solution
for real $z$ satisfying $0<z<2$, so this is the only minimal solution we need.

@@

With ${_2F_1}$ computed, the only remaining numerical danger in computing
$p(m)$ in~\eq{eq:p_m_bursty+rep_appdx} is overflow of the gamma functions.
This is easily solved by taking the log of the entire expression
and using standard routines to compute the log of the gamma functions,
then exponentiating the entire expression at the end if $p(m)$
is needed rather than $\log p(m)$.

### 2020-05-23 doc/section_04_bayesian_inference.tex
Text on posterior inference on repressed data

@@

\mmnote{Outline:
\begin{itemize}
\item Full model. Write Bayes, handwave 2F1 dist story in limits of weak,
strong, and intermediate repression.
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}

>>>>

Now that we have a satisfactory model in hand for constitutive
promoters, we would like to return to the main thread: can we
infer repressor binding and unbinding rates from mRNA distributions
over a population of cells as measured by FISH in~\cite{Jones2014}?
If so, how do these inferred rates compare to direct
single-molecule measurements such as from~\cite{Hammar2014} and
to binding energies such as from~\cite{Garcia2011a}
and~\cite{Razo-Mejia2018}, which were inferred under the assumptions
of the equilibrium models in Figure~\ref{fig1:means_cartoons}(B)?
And can this comparison shed light on the unreasonable effectiveness
of the equilibrium models?

@@

As we found in Section~\ref{sec:beyond_means}, for our purposes
the ``right'' model of a constitutive promoter is the bursty
picture, model five in Figure~\ref{fig:constit_cartoons}.
Therefore our starting point here is the analogous model with
repressor added, model 5 in Figure~\ref{fig1:means_cartoons}.
For a given repressor binding site and copy number,
this model has four rate parameters to be inferred:
the repressor binding and unbinding rates $k_R^+$, and $k_R^-$,
the initiation rate of bursts, $k_i$, and the mean burst size $b$
(we nondimensionalize all of these by the mRNA degradation rate $\gamma$).

Before showing the mathematical formulation of our statistical
inference model, we would like to sketch the intuitive structure.
The dataset from~\cite{Jones2014} we consider consists of
single-cell mRNA FISH measurements of nine different conditions,
spanning several combinations of three unique binding sites and
four unique repressor copy numbers. We assume that the values of
$k_i$ and $b$ are known, since we have already cleanly inferred
them from constitutive promoter data, and further we assume that
these values are the same across datasets with different
repressor binding sites and copy numbers. We assume that there is
one unbinding rate parameter for each repressor binding site,
which is shared across conditions, and likewise one binding rate
for each unique repressor copy number. This makes our model seven
dimensional, or nine if one counts $k_i$ and $b$ as well.

Formally now, denote the set of seven repressor rates to be inferred as
\begin{equation}
\vect{k} =\{k_{Oid}^-, k_{O1}^-, k_{O2}^-,
k_{0.5}^+, k_{1}^+, k_{2}^+, k_{10}^+\}.
\end{equation}
Note the repressor copy numbers are not known directly, so we label their
association rates by the concentration of a small molecule inducer aTc
which indirectly controls the transcription of $\textit{lacI}$.
Also note that the authors of~\cite{Jones2014} report estimates
of LacI copy number per cell rather than direct measurements.
However, these estimates were made \textit{assuming} the validity
of the equilibrium models in Figure~\ref{fig1:means_cartoons},
and since \textit{testing} these models is our goal, we will make
no assumptions about the LacI copy number for given aTc concentrations.

Bayes theorem reads simply
\begin{equation}
p(\vect{k}, k_i, b \mid D)
\propto
p(D \mid\vect{k}, k_i, b) p(\vect{k}, k_i, b),
\end{equation}
where $D$ is the set of all $N$ observed single-cell mRNA counts
across the various conditions. We assume that individual single-cell
measurements are independent so that the likelihood factorizes as
\begin{equation}
p(D \mid\vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid \vect{k}, k_i, b)
= \prod_{j=1}^N p(m\mid k_j^+, k_j^-, k_i, b)
\end{equation}
where $k_j^\pm$ represent the appropriate binding and unbinding
rates for the $j$-th measured cell. The probability
$p(m\mid k_j^+, k_j^-, k_i, b)$ appearing in the last expression
is exactly Eq.~\ref{eq:p_m_bursty+rep_appdx}, the steady-state
distribution for our bursty model with repression derived in
Section~\ref{sec:gen_fcn_appdx}, which for completeness we reproduce here as
\begin{equation}
\begin{split}
p(m \mid k_R^+, k_R^-, k_i, b) = & ~\frac{
        \Gamma(\alpha + m)\Gamma(\beta + m)\Gamma(k_R^+ + k_R^-)
        }
        {
        \Gamma(\alpha)\Gamma(\beta)\Gamma(k_R^+ + k_R^- + m)
        }
\frac{b^m}{m!}
\\
&\times {_2F_1}(\alpha+m, \beta+m, k_R^++k_R^-+m; -b).
\end{split}
\label{eq:p_m_bursty+rep}
\end{equation}
where $\alpha$ and $\beta$, defined for notational convenience, are
\begin{align}
\begin{split}
\alpha &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ + \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right)
\\
\beta &= \frac{1}{2}
\left(k_i+k_R^-+k_R^+ - \sqrt{(k_i+k_R^-+k_R^+)^2 - 4k_i k_R^-}\right).
\end{split}
\end{align}

\mmnote{Probably SI most of this paragraph?}
This likelihood is rather inscrutable. We did not find any of the
known analytical approximations for ${_2F_1}$ terribly useful in
gaining intuition, so we instead resorted to numerics. One insight we found
\mmnote{a plot would probably help explain, maybe SI, need to explain better}
was that for very strong or very weak repression, the distribution in
Eq.~\ref{eq:p_m_bursty+rep} is well approximated by a negative binomial
with burst size $b$ and burst rate $k_i$ equal to their
constitutive \textit{lacUV5} values, except with $k_i$ multiplied
by the fold-change $\left(1+k_R^+/k_R^-\right)^{-1}$.
In other words, once again only the \textit{ratio} $k_R^+/k_R^-$
was detectable. But for intermediate repression, the distribution
was visibly broadened with Fano factor greater than $1+b$, the
value for the corresponding constitutive case. This indicates
that the repressor rates had left an imprint on the distribution,
and perhaps intuitively, this intermediate regime occurs for values of
$k_R^\pm$ comparable to the burst rate $k_i$. Put another way,
if the repressor rates are much faster or much slower than $k_i$,
then there is a timescale separation and effectively only one
timescale remains, $k_i\left(1+k_R^+/k_R^-\right)^{-1}$.
Only when all three rates in the problem are comparable does the
mRNA distribution retain detectable information about them.

Next we specify priors. As for the constitutive model, weakly
informative lognormal priors are a natural choice for all our
rates. We found that if the priors were too weak, our MCMC
sampler would often become stuck in regions of parameter space
with very low probability density, unable to move. We struck a
balance in choosing our prior widths between helping the sampler
run while simultaneously verifying that the marginal posteriors
for each parameter were not artificially constrained or distorted
by the presence of the prior. The only exception to this is the
highly informative priors we placed on $k_i$ and $b$, since we
have strong knowledge of them from our inference of constitutive
promoters above.

With priors and likelihood specified we may write down our complete generative model as
\begin{equation}
\begin{split}
\log_{10}k_i &\sim \text{Normal}(0.725, 0.025)\\
\log_{10}b   &\sim \text{Normal}(0.55, 0.025)\\
\log_{10}k_{0.5}^+ &\sim \text{Normal}(-0.45, 0.3)\\
\log_{10}k_{1}^+   &\sim \text{Normal}(0.6, 0.3)\\
\log_{10}k_{2}^+   &\sim \text{Normal}(1.15, 0.3)\\
\log_{10}k_{10}^+  &\sim \text{Normal}(1.5, 0.3)\\
\log_{10}k_{Oid}^- &\sim \text{Normal}(-0.25, 0.3)\\
\log_{10}k_{O1}^-  &\sim \text{Normal}(0.1, 0.3)\\
\log_{10}k_{O2}^-  &\sim \text{Normal}(0.45, 0.3)\\
m &\sim \text{Likelihood}(k_R^+, k_R^-, k_i, b),
\end{split}
\end{equation}
where the likelihood is specified by Eq.~\ref{eq:p_m_bursty+rep}.
\mmnote{Now that I've typed this out, this notation seems totally
pointless here, especially when the likelihood is not a standard
distribution. Maybe I should make a table or something to record
the prior values, and maybe even that should go to SI?}
We ran MCMC sampling on the full nine dimensional posterior
specified by this generative model. To attempt to visualize this
object, in Figure~\ref{fig4:repressed_post_full}(A) we plot
several two-dimensional slices as contour plots, analogous to
Figure~\ref{fig:constit_post_full}(C). Each of these nine slices
corresponds to the $(k_R^+, k_R^-)$ pair of rates for one of the
conditions from the dataset used to fit the model and gives a
sense of the uncertainty and correlations in the posterior.

\mmnote{Outline:
\begin{itemize}
\item Hey look it mostly works. With full model, 9D model is well identified,
though most of the individual experiments unsurprisingly are not:
only 3 of the 9 datasets were identifiable if fit in isolation
and for the rest only the ratio of repressor rates is identifiable.
Only by making assumptions that certain rates
are equal across datasets was identifiability achieved.
PPC is not perfect, it kinda misses some
rep/op pairs but overall it's surprisingly predictive considering how strong
were the assumptions that went into it.
\item Do we chalk this up to experimental imperfections or is there interesting
biology to be found in the disagreements? I'm inclined towards the former. Even
just catching cells in truly apples-to-apples growth phases, or matching up
parameters of imaging sessions weeks or months apart, is hardly trivial.
\item The alternative is to refine our model and/or add on complexity, and it's
not at all clear what additions to make to the model, especially how to produce
the long tail of repressed strains over UV5. I see no way to get that without
getting into complicated time-correlation/history effects, where
transcription...rebounds?? After being repressed?? Sounds odd.
\item Comparison between my inferred rates, equilibrium binding E, and
single-molecule measurements. Not quite as slam-dunk as I'd hoped because of
$\gamma$ dependence, but rates are w/in factor of 2 or better, seems quite good
really! With exception of one Oid data data point, free energies are w/in about
0.5 kT also. That's not massively larger than the repeatability threshold of
fold-change measurements, at maybe 0.1 to 0.3 kT.
\item Surprising that so many model parameters can be inferred from just a few
distributions that, by eye, don't have strong features! Does require some
assumptions, e.g., about equivalence of rates across experiments, but that seems
quite reasonable.
\end{itemize}
}
